Compositional Certified Resource Bounds

### 2015-04-13 paper/main.tex
flush last page, add names to TR
avoid repetitions
typos

\usepackage[utf8]{inputenc} %for utf8 input
\usepackage{amssymb} %for shift symbol
\usepackage{amsmath}
\usepackage{listings} %for code
\lstset{moredelim=[is][\color{blue}]{@}{@},keepspaces=true}
\usepackage{mathpartir} %for typing rules
\usepackage{microtype} %better micro typing
\usepackage{stmaryrd} %for llbracket
\usepackage{mathabx} % for boxes
\usepackage{graphicx} %to include png images
\usepackage{xcolor} %for colors
\usepackage{url}
\usepackage{enumitem}
\usepackage{array} %for stupid tables
%\usepackage[firstpage]{draftwatermark}
%\SetWatermarkText{\hspace*{8in}\raisebox{5.5in}{\includegraphics[scale=0.1]{fig/aec-badge-pldi.pdf}}}
%\SetWatermarkAngle{0}

@@

Tools in the second category originate form the {\em potential method}
of amortized analysis and type systems for functional
programs~\cite{Jost03,HoffmannAH12}. It has been shown that class
definitions of object-oriented programs~\cite{Jost06} and
data-structure predicates of separation logic~\cite{Atkey10} can play
the role of the type system in imperative programs.
However, a major weakness of existing potential-based techniques
is that they can only associate \emph{potential}
with individual program variable or data structure.
For C programs, this
fails for loops as simple as \lstinline{for(i=x;i<y;i++)} where
$y-i$ decreases, but not $|i|$. %  Mutation is yet another problem.
% While functional programs relate multiple fixed inputs (the typing
% context) with one output (the expression typed), imperative programs
% can change the value of variables in scope, so the potential of these
% variables has to be updated after mutations to ensure soundness.

@@

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which we increment $x$ by a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the one of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  In the figure, we separate
logical assertions from potential functions by semicolons. Note that the
logical assertions are only used in the rule for the assignment
\code{x=x+K}.

@@

To find resource bounds automatically, we first need to restrict our
search space.  In this work, we focus on a the following form of
potential functions, which can express tight bounds for many typical
programs and allows for inference with \emph{linear programming}.
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state \land x \neq y}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
$$
%
Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
is a simplified program state that maps variable names to integers,
$|\inter a b| = \max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
references to the linear coefficients $q_i$, we introduce an \emph{index
  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
\in \Var \land x \neq y \}$.  Each index $i$ corresponds to a \emph{base function}
$f_i$ in the potential function: $0$ corresponds to the constant
function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
rewrite the above equality as\iffull{
$$ \Phi(\state) = \sum_{i \in I} q_i f_i(\state).$$}{
$\Phi(\state) = \sum_{i \in I} q_i f_i(\state)$.}%
We often write $xy$ to denote the
index $(x,y)$.  \iffull{The family $(f_i)_{i\in I}$ is actually a basis (in
the linear-algebra sense) for potential functions.}{}  This allows us to
uniquely represent any linear potential function
$\Phi$ as a \emph{quantitative annotation} $Q = (q_i)_{i \in I}$, that
is, a family of \ifshort{non-negative}{} rational numbers where only
a finite number of elements are not zero. \iffull{ For soundness, we
  require that potential functions always give non-negative results.
  This is achieved by restricting coefficients in a quantitative
  annotation to be always non-negative.}{}

@@

In addition to the quantitative annotations our automatic amortized
analysis needs to maintain a minimal abstract state to justify certain
operations on quantitative annotations.  For example when analyzing
the code $x \gets x + y$, it is helpful to know the sign of $y$ to
determine which intervals will increase or decrease.  The knowledge
needed by our rules can be inferred by local reasoning (i.e. in
basic blocks without recursion and loops) within usual theories
(e.g. Presburger arithmetic or bit vectors).

The abstract program state is represented as \emph{logical contexts} in the
derivation system used by our automated tool.
Our implementation finds these logical contexts using abstract
interpretation with the domain of linear inequalities.  We observed that
the rules of the analysis often requires only minimal local knowledge.
This means that it is not necessary for us to compute precise loop
invariants and only a rough fixpoint (e.g. keeping only inequalities on
variables unchanged by the loop) is sufficient to obtain good bounds.

@@

Example \emph{t62} (\pref{fig:ex_list_2})
is the inner loop of a quick sort implementation in cBench.  More
precisely, it is the partitioning part of the algorithm.  This partition
loop has linear complexity, and feeding it to our analysis gives a
worst-case bound $2+3|[l,h]|$.  The bound is not optimal but it can be
refined by rewriting the program. % to let our logical inference
%discover more invariants.
%With the potential
%method in mind, it is clear that the three loops get their
%potential from the same interval $[l,h]$ regardless of their nesting.
With the potential method in mind, we can reason as follows.
%
If $h \geq l$ initially, the cost of the loop is 2.
Otherwise, the cost of each round (at
most 3) is payed using the potential of $[l,h]$ by the first
increment to $l$ because we know that $l<h$.  The two inner loops
can also use $[l,h]$ to pay for their inner costs.
%
KoAT fails to
find a bound and LOOPUS derives the quadratic bound
$(h-l-1)^2$.  Following the classical technique, these tools try to find one
ranking function for each loop and combine them multiplicatively
or additively.

@@

We present the rules for a subset of the
expressions of Clight.  Clight is the first intermediate
language of the CompCert compiler~\cite{CompcertCACM}, it
is a cleaner C with a unified looping construct and
side-effect free expressions.  We reuse most of CompCert's
syntax but instrument the semantics with a resource metric
$M$ that accounts for the cost (an arbitrary rational number)
of each step in the operational semantics.  For example, $M_e(\code{exp})$ is the cost of
evaluating the expression \code{exp}; the rationals $M_f$
and $M_r$ account respectively for the cost of a call to the
function $f$ and the cost of returning from it.
More details are provided in \pref{sec:soundness}.

@@

The derivation system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The derivation rules derive judgements of the form
$$
  (\Gamma_B; Q_B), (\Gamma_R; Q_R) \vdash
    \htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }.
$$
%
The part $\htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }$ of the judgement
can be seen as a quantitative Hoare triple.  All assertions are split into
two parts, the logical part and the quantitative part.  The
quantitative part $Q$ represents a potential function as a collection
of non-negative numbers $q_i$ indexed by the index set $I$.  The
logical part $\Gamma$ is left abstract but is enforced by our
derivation system to respect classic Hoare logic constraints.
%
The meaning of this basic judgment is as follows: If $S$ is executed
with starting state $\state$, the assertions in $\Gamma$
hold, and at least $Q(\state)$ resources are available then the
evaluation does not run out of resources and, if the execution
terminates in state $\state'$, there are at least $Q(\state')$
resources left and $\Gamma'$ holds for $\state'$.

@@

The rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc Q:Inc}
describe how the potential is distributed after a size change of
a variable.
\iffull{The rule {\sc Q:IncP} is for increments $x \gets x + y$
and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
that can be applied if $y$ is not positive.}{The rule {\sc Q:IncP} is for increments
$x \gets x+y$, it applies when we can deduce from the logical context
$\Gamma$ that $y \ge 0$.  Of course, we have symmetrical rules that
can be applied for decrements and when $y$ is not positive.}
The rules are all equivalent in the case where $y=0$.
The rule {\sc Q:Inc} can be applied
if we cannot find the sign of $y$.  % In $\toolname$,
% the rules for increments and decrements are combined into one rule\iffull{ where
% the specialized rules take priority over {\sc Q:Inc}.}{.}

@@

The rule {\sc Q:Weak} is not syntax directed. In the implementation
we apply {\sc Q:Weak} before loops and between the two statements of a
sequential composition. We could integrate weakening into every
syntax directed rule but this simple heuristic helps to make the analysis
efficient.  The high-level idea of {\sc Q:Weak} is the following: If
we have a sound judgement, then it is sound to add more potential to the
precondition and remove potential from the postcondition.  The concept of
\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
that is defined in the rule {\sc Relax} that
also deals with the important task of transferring constant potential
(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
from the logical context that the interval size $|\inter x y| \geq l$ is
larger than a constant $l$ then we can turn the potential $q_{xy} {\cdot}
|\inter x y|$ form the interval into the constant potential $l{\cdot}q_{xy}$
and guarantee that we do not gain potential.  Conversely, if $|\inter
x y| \leq u$ for a constant $u$ then we can transfer constant potential
$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
without gaining potential.

@@

We separate the search of a derivation into two steps.  As a first
step we go through the functions of the program and apply inductively
the derivation rules of the automatic amortized analysis.  This is
done in a bottom-up way for each strongly connected component (SCC) of the
call graph.  During this
process our tool uses symbolic names for the rational coefficients
$q_i$ in the rules.  Each time a linear constraint must be
satisfied by these coefficients, it is recorded in a global list for the SCC using
the symbolic names.  We reuse the constraint list for
every call from outside the SCC.

We then feed the collected constraints to an
off-the-shelf LP solver (currently CLP~\cite{clp}).
If the solver successfully finds a solution, we know that a derivation
exists and extract the values for the initial
$Q$ from the solver to get a resource bound for the program.  To get
a full derivation we extract the complete solution from the
solver, and apply it to the symbolic names $q_i$ of the coefficients
in the derivation.  If the LP solver fails to find a solution, an
error is reported.

@@

More precisely, the technique consists of the following steps. First,
a program $P$ that fails to be analyzed automatically is enriched by
auxiliary variables $\vec x$ and assertions to form a program
$P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$ for the
logical variables is selected to satisfy the proposition:
$$
\forall n\, \sigma\, \sigma'.\,
  (\sigma, P_l(\vec X(\sigma))) \Downarrow_n \sigma'
    {\implies}
  \exists n' {\le} n.\,
    (\sigma, P) \Downarrow_{n'} \sigma'.
$$
Since the annotated program and the original one are usually
syntactically close, the proof of this result goes
by simple induction on the resource-aware evaluation judgement.
Third, using existing automation tools, a bound $B(\vec x)$ for
$P_l(\vec x)$ is derived.  Finally this bound, instantiated with
$\vec X$, gives the final resource bound for the program $P$.

@@

Generally, we observed that because the instrumented program
is structurally same as the original one, it is enough to prove
that the added assertions never fail in order to show the two programs
have identical resource consumption.  This can usually be piggybacked
on standard static-analysis tools.

@@

We define quantitative Hoare triples
as $B;R\vdash_L \htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
maps from program states to an element of $\Qplusz\cup\{\infty\}$ that represents
an amount of resources available.  The assertions $B$ and $R$ are
postconditions for the case in which the block $S$ exits by a \code{break}
or \code{return} statement.  Additionally, $R$ depends on the return
value of the current function.  The meaning of the triple $\htriple Q
S {Q'}$ is as follows: If $S$ is executed with starting state $\state$,
the empty continuation, and at least $Q(\state)$ resource units available
then the evaluation does not run out of resources and there are at
least $Q(\state')$ resources left if the evaluation terminates in
$\state'$.  The logic rules are similar to the ones in previous
work and generalized to account for the cost introduced by  our
cost-aware semantics.

Finally, we define a strong compositional continuation based soundness for
triples and prove the validity of all the rules in Coq.  \ifshort{The full
version of this paper \cite{anon_extended}, provides explanations for
the rules and a thorough overview of our soundness proof.}{Details about
the logic and its soundness proof are provided in the \pref{app:logic}.}

@@

The translation of an assertion $(\Gamma, Q)$ in the automatic
analysis is defined by
$$
  \tr(\Gamma, Q) := \lambda\state .\, \Gamma(\state) + \Phi_Q(\state),
$$
where we write $\Phi_Q$ for the unique linear potential function
defined by the quantitative annotation $Q$.  The logical context $\Gamma$
is implicitly lifted to a quantitative assertion by mapping a state
$\state$ to 0 if $\Gamma(\state)$ holds and to $\infty$ otherwise.
\iffull{We also need to
translate the assumptions in function contexts of the automatic
analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=
\forall z \, \vec v \, v . $
%
$$
(
  \lambda \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]),
  \lambda v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret {\mapsto} v])
  )
$$}{}
%
These definitions let us translate the judgement
$J := B, R \vdash \htriple P S {P'}$\iffull{ in the context $\Delta$ by $\tr(J) :=$}{ by}
\iffull{$$
  (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash
    \htriple{\tr(P)}{S}{\tr(P')}.
$$}{%
$$
  \tr(J) := \tr(B); \tr(R) \vdash_L \htriple{\tr(P)} S {\tr(P')}.
$$}%
\noindent{}The soundness of the automatic analysis can now be stated
formally with the following theorem.
%
\begin{theorem}[Soundness of the automatic analysis]
  If $J$ is a judgement derived by the automatic analysis,
  then $\tr(J)$ is a quantitative Hoare triple derivable
  in the quantitative logic.
\end{theorem}
%
\noindent
The proof of this theorem is constructive and maps each rule
of the automatic analysis directly to its counterpart in the
quantitative logic.  The most tricky parts are the translations of the
rules for increments and decrements and the rule {\sc Q:Weak} for
weakening because they make essential use of the algebraic properties
of our linear potential functions.
\iffull{ For instance we prove the following lemma to show the soundness
of the translation of {\sc Q:Weak}.

@@

\pref{fig:compar} shows five representative loop patterns from the
evaluation.
Example \emph{t09} is a loop that performs an expensive operation every
4 steps.  $\toolname$ is the only tool able to amortize this cost over
the input parameter $x$.
Example \emph{t19}
demonstrates the compositionality of the analysis.  The program consists
of two loops that decrement a variable $i$.  In the first loop, $i$ is
decremented down to 100 and in the second loop $i$ is decremented further
down to $-1$.  However, between the loops we assign \code{i=i+k+50}.
So in total the program performs $52 + |[-1,i]| + |[0,k]|$ ticks.
Our analysis finds this tight bound because our amortized analysis
naturally takes into account the relation between the two loops.
Example \emph{t30} decrements both input variables $x$ and $y$ down to
zero in an unconventional way.  In the loop body, first $x$ is decremented
by one.  Then the values of the variables $x$ and $y$ are switched using
the local variable $t$ as a buffer.  Our analysis infers the tight
bound $|[0,x]|+|[0,y]|$.  Sometimes we need some assumptions on the
inputs in order to derive a bound.  Example \emph{t15} is such a case.
We assume here that the input variable $y$ is non-negative and write
\lstinline{assert(y>=0)}.  The assignment \lstinline{x -= y+1} in the
loop is split in \lstinline{x--} and \lstinline{x -= y}. If we enter the
loop then we know that $x>0$, so we can obtain constant potential from
\lstinline{x--}.  Then we know that $x\geq y\geq 0$, as a consequence
we can share the potential of $|[0,x]|$ between $|[0,x]|$ and $|[0,y]|$
after \lstinline{x -= y}.  % In this way, we derive a tight linear bound.

@@

Our current implementation of $\toolname$ does not currently support
all of Clight.  Programs with function pointers, goto statements,
continue statements, and dereferences of stack-allocated variables
cannot be analyzed automatically.  While these limitations concern the
current implementation, our technique is in principle capable to handle
them.

>>>>

\usepackage[utf8]{inputenc} %for utf8 input
\usepackage{amssymb} %for shift symbol
\usepackage{amsmath}
\usepackage{listings} %for code
\lstset{moredelim=[is][\color{blue}]{@}{@},keepspaces=true}
\usepackage{mathpartir} %for typing rules
\usepackage{microtype} %better micro typing
\usepackage{stmaryrd} %for llbracket
\usepackage{mathabx} % for boxes
\usepackage{graphicx} %to include png images
\usepackage{xcolor} %for colors
\usepackage{url}
\usepackage{enumitem}
\usepackage{array} %for stupid tables
\usepackage{flushend} %equalize columns length on last page
%\usepackage[firstpage]{draftwatermark}
%\SetWatermarkText{\hspace*{8in}\raisebox{5.5in}{\includegraphics[scale=0.1]{fig/aec-badge-pldi.pdf}}}
%\SetWatermarkAngle{0}

@@

Tools in the second category originate form the {\em potential method}
of amortized analysis and type systems for functional
programs~\cite{Jost03,HoffmannAH12}. It has been shown that class
definitions of object-oriented programs~\cite{Jost06} and
data-structure predicates of separation logic~\cite{Atkey10} can play
the role of the type system in imperative programs.
However, a major weakness of existing potential-based techniques
is that they can only associate \emph{potential}
with individual program variables or data structures.
For C programs, this
fails for loops as simple as \lstinline{for(i=x;i<y;i++)} where
$y-i$ decreases, but not $|i|$. %  Mutation is yet another problem.
% While functional programs relate multiple fixed inputs (the typing
% context) with one output (the expression typed), imperative programs
% can change the value of variables in scope, so the potential of these
% variables has to be updated after mutations to ensure soundness.

@@

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which we increment $x$ by a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the one of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  In the figure, we separate
logical assertions from potential functions with semicolons. Note that the
logical assertions are only used in the rule for the assignment
\code{x=x+K}.

@@

To find resource bounds automatically, we first need to restrict our
search space.  In this work, we focus on the following form of
potential functions, which can express tight bounds for many typical
programs and allows for inference with \emph{linear programming}.
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state \land x \neq y}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
$$
%
Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
is a simplified program state that maps variable names to integers,
$|\inter a b| = \max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
references to the linear coefficients $q_i$, we introduce an \emph{index
  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
\in \Var \land x \neq y \}$.  Each index $i$ corresponds to a \emph{base function}
$f_i$ in the potential function: $0$ corresponds to the constant
function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
rewrite the above equality as\iffull{
$$ \Phi(\state) = \sum_{i \in I} q_i f_i(\state).$$}{
$\Phi(\state) = \sum_{i \in I} q_i f_i(\state)$.}%
We often write $xy$ to denote the
index $(x,y)$.  \iffull{The family $(f_i)_{i\in I}$ is actually a basis (in
the linear-algebra sense) for potential functions.}{}  This allows us to
uniquely represent any linear potential function
$\Phi$ as a \emph{quantitative annotation} $Q = (q_i)_{i \in I}$, that
is, a family of \ifshort{non-negative}{} rational numbers where only
a finite number of elements are not zero. \iffull{ For soundness, we
  require that potential functions always give non-negative results.
  This is achieved by restricting coefficients in a quantitative
  annotation to be always non-negative.}{}

@@

In addition to the quantitative annotations, our automatic amortized
analysis needs to maintain a minimal abstract state to justify certain
operations on quantitative annotations.  For example when analyzing
the code $x \gets x + y$, it is helpful to know the sign of $y$ to
determine which intervals will increase or decrease.  The knowledge
needed by our rules can be inferred by local reasoning (i.e. in
basic blocks without recursion and loops) within usual theories
(e.g. Presburger arithmetic or bit vectors).

The abstract program state is represented as \emph{logical contexts} in the
derivation system used by our automated tool.
Our implementation finds these logical contexts using abstract
interpretation with the domain of linear inequalities.  We observed that
the rules of the analysis often require only minimal local knowledge.
This means that it is not necessary for us to compute precise loop
invariants and only a rough fixpoint (e.g. keeping only inequalities on
variables unchanged by the loop) is sufficient to obtain good bounds.

@@

Example \emph{t62} (\pref{fig:ex_list_2})
is the inner loop of a quick sort implementation in cBench.  More
precisely, it is the partitioning part of the algorithm.  This partition
loop has linear complexity, and feeding it to our analysis gives the
worst-case bound $2+3|[l,h]|$.  This bound is not optimal but it can be
refined by rewriting the program. % to let our logical inference
%discover more invariants.
%With the potential
%method in mind, it is clear that the three loops get their
%potential from the same interval $[l,h]$ regardless of their nesting.
To understand the bound, we can reason as follows.
%
If $h \geq l$ initially, the cost of the loop is 2.
Otherwise, the cost of each round (at
most 3) can be payed using the potential of $[l,h]$ by the first
increment to $l$ because we know that $l<h$.  The two inner loops
can also use $[l,h]$ to pay for their inner costs.
%
KoAT fails to
find a bound and LOOPUS derives the quadratic bound
$(h-l-1)^2$.  Following the classical technique, these tools try to find one
ranking function for each loop and combine them multiplicatively
or additively.

@@

We present the rules for a subset of Clight.
Clight is the first intermediate
language of the CompCert compiler~\cite{CompcertCACM}, it
is a cleaner C with a unified looping construct and
side-effect free expressions.  We reuse most of CompCert's
syntax but instrument the semantics with a resource metric
$M$ that accounts for the cost (an arbitrary rational number)
of each step in the operational semantics.  For example, $M_e(\code{exp})$ is the cost of
evaluating the expression \code{exp}; the rationals $M_f$
and $M_r$ account respectively for the cost of a call to the
function $f$ and the cost of returning from it.
More details are provided in \pref{sec:soundness}.

@@

The derivation system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The derivation rules derive judgements of the form
$$
  (\Gamma_B; Q_B), (\Gamma_R; Q_R) \vdash
    \htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }.
$$
%
The part $\htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }$ of the judgement
can be seen as a quantitative Hoare triple.  All assertions are split into
two parts, the logical part and the quantitative part.  The
quantitative part $Q$ represents a potential function as a collection
of non-negative numbers $q_i$ indexed by the index set $I$.  The
logical part $\Gamma$ is left abstract but is enforced by our
derivation system to respect classic Hoare logic constraints.
%
The meaning of this basic judgment is as follows: If $S$ is executed
with starting state $\state$, the assertions in $\Gamma$
hold, and at least $Q(\state)$ resources are available then the
evaluation does not run out of resources and, if the execution
terminates in state $\state'$, there are at least $Q'(\state')$
resources left and $\Gamma'$ holds for $\state'$.

@@

The rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc Q:Inc}
describe how the potential is distributed after a size change of
a variable.
The rule {\sc Q:IncP} is for increments $x \gets x + y$
and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
that can be applied if $y$ is not positive.
The rules are all equivalent in the case where $y=0$.
The rule {\sc Q:Inc} can be applied
if we cannot find the sign of $y$.  % In $\toolname$,
% the rules for increments and decrements are combined into one rule\iffull{ where
% the specialized rules take priority over {\sc Q:Inc}.}{.}

@@

The rule {\sc Q:Weak} is the only rule not syntax directed. %In the implementation
%we apply {\sc Q:Weak} before loops and between the two statements of a
%sequential composition.
We could integrate weakenings into every syntax directed rule but, for
the sake of efficiency, the implementation uses a simple heuristic instead.
The high-level idea of {\sc Q:Weak} is the following: If
we have a sound judgement, then it is sound to add more potential to the
precondition and remove potential from the postcondition.  The concept of
\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
that is defined in the rule {\sc Relax}. This rule
also deals with the important task of transferring constant potential
(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
from the logical context that the interval size $|\inter x y| \geq l$ is
larger than a constant $l$ then we can turn the potential $q_{xy} {\cdot}
|\inter x y|$ form the interval into the constant potential $l{\cdot}q_{xy}$
and guarantee that we do not gain potential.  Conversely, if $|\inter
x y| \leq u$ for a constant $u$ then we can transfer constant potential
$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
without gaining potential.

@@

We separate the search of a derivation in two steps.  As a first
step we go through the functions of the program and apply inductively
the derivation rules of the automatic amortized analysis.  This is
done in a bottom-up way for each strongly connected component (SCC) of the
call graph.  During this
process our tool uses symbolic names for the rational coefficients
$q_i$ in the rules.  Each time a linear constraint must be
satisfied by these coefficients, it is recorded in a global list for the SCC using
the symbolic names.  We reuse the constraint list for
every call from outside the SCC.

We then feed the collected constraints to an
off-the-shelf LP solver (currently CLP~\cite{clp}).
If the solver successfully finds a solution, we know that a derivation
exists and extract the values for the initial
$Q$ from the solver to get a resource bound for the program.  To get
a full derivation, we extract the complete solution from the
solver and apply it to the symbolic names $q_i$ of the coefficients
in the derivation.  If the LP solver fails to find a solution, an
error is reported.

@@

More precisely, the technique consists of the following steps. First,
a program $P$ that fails to be analyzed automatically is enriched by
auxiliary variables $\vec x$ and assertions to form a program
$P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$ for the
logical variables is selected to satisfy the proposition:
\begin{equation} \tag{*}
\forall n\, \sigma\, \sigma'.\,
  (\sigma, P_l(\vec X(\sigma))) \Downarrow_n \sigma'
    {\implies}
  \exists n' {\le} n.\,
    (\sigma, P) \Downarrow_{n'} \sigma'.
\end{equation}
Since the annotated program and the original one are usually
syntactically close, the proof of this result goes
by simple induction on the resource-aware evaluation judgement.
Third, using existing automation tools, a bound $B(\vec x)$ for
$P_l(\vec x)$ is derived.  Finally this bound, instantiated with
$\vec X$, gives the final resource bound for the program $P$.

@@

Generally, we observed that because the instrumented program
is structurally same as the original one, it is enough to prove
that the added assertions never fail in order to show the two programs
satisfy the proposition (*).  This can usually be piggybacked
on standard static-analysis tools.

@@

We define quantitative Hoare triples
as $B;R\vdash_L \htriple Q S {Q'}$ where $B$, $R$, $Q$, and $Q'$ are
maps from program states to an element of $\Qplusz\cup\{\infty\}$ that represents
an amount of resources available.  The assertions $B$ and $R$ are
postconditions for the case in which the block $S$ exits by a \code{break}
or \code{return} statement.  Additionally, $R$ depends on the return
value of the current function.  The meaning of the triple $\htriple Q
S {Q'}$ is as follows: If $S$ is executed with starting state $\state$,
the empty continuation, and at least $Q(\state)$ resource units available
then the evaluation does not run out of resources and there are at
least $Q'(\state')$ resources left if the evaluation terminates in
$\state'$.  The logic rules are similar to the ones in previous
work and generalized to account for the cost introduced by  our
cost-aware semantics.

Finally, we define a strong compositional continuation-based soundness for
triples and prove the validity of all the rules in Coq.  \ifshort{The full
version of this paper \cite{anon_extended}, provides explanations for
the rules and a thorough overview of our soundness proof.}{Details about
the logic and its soundness proof are provided in the \pref{app:logic}.}

@@

The translation of an assertion $(\Gamma; Q)$ in the automatic
analysis is defined by
$$
  \tr(\Gamma; Q) := \lambda\state .\, \Gamma(\state) + \Phi_Q(\state),
$$
where we write $\Phi_Q$ for the unique linear potential function
defined by the quantitative annotation $Q$.  The logical context $\Gamma$
is implicitly lifted to a quantitative assertion by mapping a state
$\state$ to 0 if $\Gamma(\state)$ holds and to $\infty$ otherwise.
\iffull{We also need to
translate the assumptions in function contexts of the automatic
analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=
\forall z \, \vec v \, v . $
%
$$
(
  \lambda \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]),
  \lambda v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret {\mapsto} v])
  )
$$}{}
%
These definitions let us translate the judgement
$J := B, R \vdash \htriple P S {P'}$\iffull{ in the context $\Delta$ by $\tr(J) :=$}{ by}
\iffull{$$
  (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash
    \htriple{\tr(P)}{S}{\tr(P')}.
$$}{%
$$
  \tr(J) := \tr(B); \tr(R) \vdash_L \htriple{\tr(P)} S {\tr(P')}.
$$}%
\noindent{}The soundness of the automatic analysis can now be stated
formally with the following theorem.
%
\begin{theorem}[Soundness of the automatic analysis]
  If $J$ is a judgement derived by the automatic analysis,
  then $\tr(J)$ is a quantitative Hoare triple derivable
  in the quantitative logic.
\end{theorem}
%
\noindent
The proof of this theorem is constructive and maps each rule
of the automatic analysis directly to its counterpart in the
quantitative logic.  The most tricky parts are the translations of the
rules for increments and decrements and the rule {\sc Q:Weak} for
weakening because they make essential use of the algebraic properties
of our linear potential functions.
\iffull{ For instance we prove the following lemma to show the soundness
of the translation of {\sc Q:Weak}.

@@

\pref{fig:compar} shows five representative loop patterns from the
evaluation.
Example \emph{t09} is a loop that performs an expensive operation every
4 steps.  $\toolname$ is the only tool able to amortize this cost over
the input parameter $x$.
Example \emph{t19}
demonstrates the compositionality of the analysis.  The program consists
of two loops that decrement a variable $i$.  In the first loop, $i$ is
decremented down to 100 and in the second loop $i$ is decremented further
down to $-1$.  However, between the loops we assign \lstinline{i += k+50}.
So in total the program performs $52 + |[-1,i]| + |[0,k]|$ ticks.
Our analysis finds this tight bound because our amortized analysis
naturally takes into account the relation between the two loops.
Example \emph{t30} decrements both input variables $x$ and $y$ down to
zero in an unconventional way.  In the loop body, first $x$ is decremented
by one, then the values of the variables $x$ and $y$ are switched using
the local variable $t$ as a buffer.  Our analysis infers the tight
bound $|[0,x]|+|[0,y]|$.  Sometimes we need some assumptions on the
inputs in order to derive a bound.  Example \emph{t15} is such a case.
We assume here that the input variable $y$ is non-negative and write
\lstinline{assert(y>=0)}.  The assignment \lstinline{x -= y+1} in the
loop is split in \lstinline{x--} and \lstinline{x -= y}. If we enter the
loop then we know that $x>0$, so we can obtain constant potential from
\lstinline{x--}.  Then we know that $x\geq y\geq 0$, as a consequence
we can share the potential of $|[0,x]|$ between $|[0,x]|$ and $|[0,y]|$
after \lstinline{x -= y}.  % In this way, we derive a tight linear bound.

@@

Our implementation does not currently support
all of Clight.  Programs with function pointers, goto statements,
continue statements, and dereferences of stack-allocated variables
cannot be analyzed automatically.  While these limitations concern the
current implementation, our technique is in principle capable to handle
them.
